### 2018-07-13 submission/manuscript.tex
fix intro with lab suggestions

\textbf{Conclusions.} Although there are 16S rRNA gene sequencing
differences based on polymerase and number of cycles used, they are
small with respect to the biological differences between fecal bacterial
communities within indivdiuals. Collectively, these results provide
evidence that a real biological difference between groups, based on 16S
rRNA gene sequencing of fecal samples, should be detectable regardless
of polymerase and number of cycles used.

@@

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Some of this variation could be
due to error introduced during the 16S rRNA gene sequencing workflow. A
typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors in the observed bacterial community composition based on the kits
used, but that these differences are smaller than the overall biological
difference measured between samples with different kits (Song et al.,
2016; Bassis et al., 2017). Since these studies use the same PCR
approach while varying preservation or extraction method, the
contribution of PCR error to this overall workflow is not well
characterized.

There is a large body of literature that shows there are errors due to
primer and number of cycles chosen for the PCR stage of 16S rRNA gene
sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
variable region dependent binding affinities which causes an inability
to detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Another source of error is the selective amplification of
AT-rich over GC-rich sequences which exaggerate the difference between
16S rRNA genes higher in AT versus those higher in GC content (Polz \&
Cavanaugh, 1998). Many of these sources of error are made worse as the
number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
Kebschull \& Zador, 2015). For example, both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) also
can increase as the number of cycles increases. This will increase the
number of Operational Taxonomic Units (OTUs) observed and drastically
change the values obtained from commonly used diversity measures (Acinas
et al., 2005; Santos et al., 2016). Additionally, as the number of
cycles increases more chimeras can form from an aborted extension step
that causes a priming error and subsequent secondary extension (Haas et
al., 2011). These chimeras will artificially increase community
diversity by increasing the number of OTUs that are observed (Haas et
al., 2011). In addition to these sources of errors, there also are
multiple families of DNA polymerases that have their own error rate and
proof reading capacity (Ishino \& Ishino, 2014). Interestingly, the
influence that these different DNA polymerases can have on the observed
16S rRNA gene sequencing results have not been well studied like some of
the other sources of PCR-based error

A recent study found differences in the number of OTUS and chimeras
between normal and high fidelity DNA polymerases (Gohl et al., 2016).
The authors of this study could reduce the difference between two
polymerases by optimizing the annealing and extension steps within the
PCR protocol (Gohl et al., 2016). Yet, within this study there was no
comparison made between different high fidelity DNA polymerases on the
same samples. Additionally, the fecal samples that were analyzed, as a
model of a more complex community, had different processing methodolgies
applied to the two different treatment groups tested, which could
exagerate the observed biological differences. Due to these gaps, it is
natural to extend the previous line of inquiry to include a more
detailed examination of two specific areas. First, for the investigation
of whether differences in PCR error are detected between high fidelity
polymerases and number of cycles used. Second, for examining whether
these differences in error are enough to obscure smaller biological
signals than what has been previously reported (Gohl et al., 2016).
Thus, this study will investigate whether differences in high fidelity
DNA polymerases are due to error and whether these differences obscure
the biological variation in the fecal bacterial community composition
between individuals. We accomplish this by examining GC-based
amplification error, sequence error (e.g.~subsitutions), chimera
prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
differences at varying number of cycles in five different high fidelity
DNA polymerases. Additionally, we also created Random Forest models to
classify number of cycles, polymerase, and individuals to assess which
model had the best classification success when using 16S rRNA gene
sequencing data.

@@

\textbf{\emph{Sequence processing.}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016). For
simplicity, the use of the term `error' within the context of this study
moving forward encompasses changes in the downstream sequencing results
due to the amplification step (e.g.~higher GC content genomes, base
subsitutions, deletions, insertions, non-specific amplification,
chimeras, etc.).

@@

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). High and low GC content was determined based on the median GC
percentage of either the V4 region or the whole genome of the bacterial
species used in the mock community. Differences in the total number of
OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
comparison of the number of OTUS in fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation. Bray-Curtis distance matrices were generated using
mothur. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
within R. The Random Forest models were run using the caret package (v
6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
different 80/20 splits of the data was run to generate a range of the
Logloss value. The probability of a correct call was obtained from this
Logloss value by taking the negative natural logrithim. For both error
and chimera analysis, samples were tested using Kruskal-Wallis with a
Dunns post-hoc test. Where applicable, correction for multiple
comparison utilized the Benjamini-Hochberg method (Benjamini \&
Hochberg, 1995).

@@

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used.}} In addition to the sequence error and chimera differences
between polymerases, we also investigated whether there were polymerase
or number of cycle dependent effects on the observed number of OTUs.
There was a polymerase dependent difference in the range of the number
of OTUs in the mock community samples {[}Figure 3A{]}. The closest any
of the polymerases came to a total of 8 OTUs, created by running the
mock reference 16S sequences through mothur processing, was at 25 and 30
cycles {[}Figure 3A{]}. Differences between polymerases for the number
of OTUs created were observed as early as 20 cycles in the mock
community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
Tukey post-hoc test, the majority of differences for the number of OTUs
detected in the mock community was largely due to Kappa and Platinum
versus the other polymerases across the number of cycles used {[}Table
S5{]}. Similarily, differences in the range of the number of OTUs
detected for fecal samples was dependent on the polymerase used
(e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion at 35 cycles = 84
- 136) {[}Figure 3B{]}. There was also a trend for lower number of
cycles (15-20) to result in a reduced range in the number of OTUs versus
higher number of cycles (25, 30, and 35) for all polymerases
(e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35 cycles = 84 -
136) {[}Figure 3B{]}. Unlike mock community samples, an overall
difference in the number of OTUs was only detected between polymerases
at 35 cycles in fecal samples (F-stat \textgreater{} 16.35, P-value =
9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to identify
any differences between specific polymerase groups (P-value
\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
noticeably larger difference between polymerases in the number of OTUs
than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
consistent across the number of amplification cycles tested {[}Figure
3{]}. Based on these observations, it is clear that larger differences
in total number of OTUs based on polymerases are found in mock versus
fecal communities.

@@

\textbf{\emph{Variation due to polymerase and number of cycles is
dwarfed by the biological variation of an individual.}} Using 16S rRNA
gene sequencing data there was a large disparity in a model's ability to
correctly classify samples based on polymerase used, number of cycles,
and individual {[}Figure 5A{]}. The model classification success for an
individual was significantly better than all other models tested
{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most impotant
variables to the polymerase and number of cycle models were not the most
important variables in the model classifying individuals {[}Figure
5B{]}. This observation shows that OTUs that change due to the number of
cycles and polymerase used are not the same as OTUs that differentiate
between indvidiuals {[}Figure 5B{]}. Although there are a variety of
errors associated with the amplification process, they are smaller than
the bacterial community variation between indviduals.

@@

Previous studies have reported GC dependent amplification differences
(Polz \& Cavanaugh, 1998) and we confirm that these differences still
persist in high fidelity DNA polymerase when analyzing mock community
samples {[}Figure 1{]}. Additionally, these specific polymerases have
different sequence error rates and chimera prevalence {[}Figure 2{]}.
Chimera prevalence in particular was strongly associated with the total
number of OTUs detected in a mock community {[}Figure 2E{]}. This total
number of OTUs also varied by polymerase used for these mock community
samples {[}Fiugre 3A{]}. The range in the number of OTUs observed
increased with the number of cycles and the change in range was
dependent on the polymerase for both mock community and fecal samples
{[}Figure 3{]}. However, these differences did not translate to large
bacterial community composition changes based on polymerase {[}Figure
4{]}. Even though there were bacterial community composition changes
based on the number of cycles they occured independently of the
polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based
differences observed within the number of cycles between any of the
polymerases for fecal samples. When Random Forest models were created to
classify number of cycles, polymerase, and individual based on the 16S
rRNA gene sequencing, the model performance for individuals was
significantly greater than all other models tested {[}Figure 5A{]}. In
addition to the better model performance of individuals, the most
important OTUs to this model were not the most important OTUs for the
other models tested {[}Figure 5B{]}. Our observations show that
differences between high fidelity DNA polymerases can be found. However,
when using 16S rRNA gene sequencing these technical differences are
dwarfed by biological differences between fecal bacterial communities
within individuals.

>>>>

\textbf{Conclusions.} Although there are 16S rRNA gene sequencing
differences based on polymerase and number of cycles used, they are
small with respect to the biological differences between fecal bacterial
communities within individuals. Collectively, these results provide
evidence that a real biological difference between groups, based on 16S
rRNA gene sequencing of fecal samples, should be detectable regardless
of polymerase and number of cycles used.

@@

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Some of this variation could be
due to error introduced during the 16S rRNA gene sequencing workflow. A
typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors in the observed bacterial community composition based on the kits
used, but that these differences are smaller than the overall biological
difference measured between samples with different kits (Song et al.,
2016; Bassis et al., 2017). Since these studies use the same PCR
approach while varying preservation or extraction method, the
contribution of PCR error to this overall workflow is not as well
characterized.

There is a large body of literature that shows there are errors due to
primer and number of cycles chosen for the PCR stage of 16S rRNA gene
sequencing (Eckert \& Kunkel, 1991; Burkardt, 2000). Primers have
variable region dependent binding affinities which causes an inability
to detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Another source of error is the selective amplification of
AT-rich over GC-rich sequences which exaggerate the difference between
16S rRNA genes higher in AT versus those higher in GC content (Polz \&
Cavanaugh, 1998). Many of these sources of error are made worse as the
number of cycles increases (Wang \& Wang, 1996; Haas et al., 2011;
Kebschull \& Zador, 2015). For example, both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) also
can increase as the number of cycles increases. This will increase the
number of Operational Taxonomic Units (OTUs) observed and drastically
change the values obtained from commonly used diversity measures (Acinas
et al., 2005; Santos et al., 2016). Additionally, as the number of
cycles increases more chimeras can form from an aborted extension step
that causes a priming error and subsequent secondary extension (Haas et
al., 2011). These chimeras will artificially increase community
diversity by increasing the number of OTUs that are observed (Haas et
al., 2011). In addition to these sources of errors, there also are
multiple families of DNA polymerases that have their own error rate and
proof reading capacity (Ishino \& Ishino, 2014). For simplicity, the use
of the term `error' within the context of this study will be used as a
broad term for changes in the downstream sequencing results due to the
amplification step (e.g.~higher GC content genomes, base substitutions,
chimeras, changes from the expected relative abundance, etc.).

Interestingly, the influence different DNA polymerases can have on the
observed 16S rRNA gene sequencing results have not been well studied
like some of the other sources of PCR-based error. A recent study found
differences in the number of OTUS and chimeras between normal and high
fidelity DNA polymerases (Gohl et al., 2016). The authors of this study
could reduce the difference between two polymerases by optimizing the
annealing and extension steps within the PCR protocol (Gohl et al.,
2016). Yet, within this study there was no comparison made between
different high fidelity DNA polymerases on the same samples. The authors
instead focused their efforts on testing specific types of amplification
methods. Additionally, the fecal samples that were analyzed, as a model
of a more complex community, may overestimate the biological variation
in human samples due to the comparison of captive versus semi-captive
red-shanked doucs. Due to these gaps, it makes sense to extend the
previous line of inquiry to include a more detailed examination of two
specific areas. First, whether differences in PCR error are detected
based on number of cycles used and high fidelity polymerases. Second,
whether these differences in error are enough to obscure smaller
biological signals than what has been previously reported (Gohl et al.,
2016).

This study will investigate whether there are error differences that are
dependent on high fidelity DNA polymerases and whether these differences
obscure the biological variation in the fecal bacterial communities
between individuals. We accomplish this by first examining GC-based
amplification error, sequence error (e.g.~substitutions), chimera
prevalence, the number of OTUs, Bray-Curtis index, and OTU-based
differences at varying number of cycles in five different high fidelity
DNA polymerases in mock communities. In addition to these metrics we
also created Random Forest models of the mock community to assess how
successful classification based on 16S rRNA gene sequencing of number of
cycles and type of polymerase could be. We then assessed if similar
differences in the number of OTUs, Bray-Curtis index, and OTUs could be
detected in human fecal samples. Additionally, we also created Random
Forest models to assess how successful they could be at classifying
fecal samples by number of cycles, polymerase used, or individual.
Collectively, our observations suggest that although detectable
differences occur based on number of cycles or high fidelity DNA
polymerase used, they are smaller than the biological differences
between individuals.

@@

\textbf{\emph{Sequence processing.}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016). For
simplicity, the use of the term `error' within the context of this study
moving forward encompasses changes in the downstream sequencing results
due to the amplification step (e.g.~higher GC content genomes, base
substitutions, deletions, insertions, non-specific amplification,
chimeras, etc.).

@@

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). High and low GC content was determined based on the median GC
percentage of either the V4 region or the whole genome of the bacterial
species used in the mock community. Differences in the total number of
OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
comparison of the number of OTUS in fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation. Bray-Curtis distance matrices were generated using
mothur. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
within R. The Random Forest models were run using the caret package (v
6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
different 80/20 splits of the data was run to generate a range of the
Logloss value. The probability of a correct call was obtained from this
Logloss value by taking the negative natural logarithm. For both error
and chimera analysis, samples were tested using Kruskal-Wallis with a
Dunns post-hoc test. Where applicable, correction for multiple
comparison utilized the Benjamini-Hochberg method (Benjamini \&
Hochberg, 1995).

@@

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used.}} In addition to the sequence error and chimera differences
between polymerases, we also investigated whether there were polymerase
or number of cycle dependent effects on the observed number of OTUs.
There was a polymerase dependent difference in the range of the number
of OTUs in the mock community samples {[}Figure 3A{]}. The closest any
of the polymerases came to a total of 8 OTUs, created by running the
mock reference 16S sequences through mothur processing, was at 25 and 30
cycles {[}Figure 3A{]}. Differences between polymerases for the number
of OTUs created were observed as early as 20 cycles in the mock
community (F-stat = 15.82, P-value = 0.002) {[}Table S4{]}. Using a
Tukey post-hoc test, the majority of differences for the number of OTUs
detected in the mock community was largely due to Kappa and Platinum
versus the other polymerases across the number of cycles used {[}Table
S5{]}. Similarly, differences in the range of the number of OTUs
detected for fecal samples was dependent on the polymerase used
(e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion at 35 cycles = 84
- 136) {[}Figure 3B{]}. There was also a trend for lower number of
cycles (15-20) to result in a reduced range in the number of OTUs versus
higher number of cycles (25, 30, and 35) for all polymerases
(e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35 cycles = 84 -
136) {[}Figure 3B{]}. Unlike mock community samples, an overall
difference in the number of OTUs was only detected between polymerases
at 35 cycles in fecal samples (F-stat \textgreater{} 16.35, P-value =
9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to identify
any differences between specific polymerase groups (P-value
\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
noticeably larger difference between polymerases in the number of OTUs
than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
consistent across the number of amplification cycles tested {[}Figure
3{]}. Based on these observations, it is clear that larger differences
in total number of OTUs based on polymerases are found in mock versus
fecal communities.

@@

\textbf{\emph{Variation due to polymerase and number of cycles is
dwarfed by the biological variation of an individual.}} Using 16S rRNA
gene sequencing data there was a large disparity in a model's ability to
correctly classify samples based on polymerase used, number of cycles,
and individual {[}Figure 5A{]}. The model classification success for an
individual was significantly better than all other models tested
{[}Figure 5A \& Table S8{]}. Additionally, the top 10 most important
variables to the polymerase and number of cycle models were not the most
important variables in the model classifying individuals {[}Figure
5B{]}. This observation shows that OTUs that change due to the number of
cycles and polymerase used are not the same as OTUs that differentiate
between individuals {[}Figure 5B{]}. Although there are a variety of
errors associated with the amplification process, they are smaller than
the bacterial community variation between individuals.

@@

Previous studies have reported GC dependent amplification differences
(Polz \& Cavanaugh, 1998) and we confirm that these differences still
persist in high fidelity DNA polymerase when analyzing mock community
samples {[}Figure 1{]}. Additionally, these specific polymerases have
different sequence error rates and chimera prevalence {[}Figure 2{]}.
Chimera prevalence in particular was strongly associated with the total
number of OTUs detected in a mock community {[}Figure 2E{]}. This total
number of OTUs also varied by polymerase used for these mock community
samples {[}Figure 3A{]}. The range in the number of OTUs observed
increased with the number of cycles and the change in range was
dependent on the polymerase for both mock community and fecal samples
{[}Figure 3{]}. However, these differences did not translate to large
bacterial community composition changes based on polymerase {[}Figure
4{]}. Even though there were bacterial community composition changes
based on the number of cycles they occurred independently of the
polymerase used {[}Figure 4{]}. Additionally, there were no OTU-based
differences observed within the number of cycles between any of the
polymerases for fecal samples. When Random Forest models were created to
classify number of cycles, polymerase, and individual based on the 16S
rRNA gene sequencing, the model performance for individuals was
significantly greater than all other models tested {[}Figure 5A{]}. In
addition to the better model performance of individuals, the most
important OTUs to this model were not the most important OTUs for the
other models tested {[}Figure 5B{]}. Our observations show that
differences between high fidelity DNA polymerases can be found. However,
when using 16S rRNA gene sequencing these technical differences are
dwarfed by biological differences between fecal bacterial communities
within individuals.
